A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-09-12 text/discussion.Rmd

One of the biggest hurdles in cancer research is the sparsity of data; ~20,000
protein-coding genes is comparable with the number of tumor samples, even with
multiple mutations per sample. We sought to
simplify the problem by employing a "knowledge-base driven analysis"
[@pathways-Atul-Butte], investigating cancer as a disease of basic cellular and
biochemical pathways, which we accomplished by translating gene-level mutations
into pathway level disruptions. Our approach differs from previously described
methods [@PATHNET] in that we chose to focus on the pathway (defined in the
methods) as the unit of disruption instead of the gene, where individual
mutations may be sufficient to alter pathway activity.

To our knowledge, this approach has not previously been attempted despite
its relative simplicity. We limited our analysis to mutations with likely
deleterious effects, including missense, nonsense, frame-shift, stop-loss,
untranslated region and splice site mutations in genes that are actively
expressed in each cancer type, thus avoiding bias from transcription coupled
repair. Our method of filtration differs from the "rank-and-cut" method of
[@PCAWG-MAIN] but represents a reasonable attempt to account for the same
biases. We also restricted our analysis to biochemical
pathways, excluding curated gene sets related to diseases, syndromes, or classes
of proteins with shared catalytic activity or conserved domains which are
potentially problematic [@pathways-Atul-Butte]. We chose this approach to limit
redundancy and exclude biologically unrelated collections of genes.

The hypothesis that cancer is the result of dysfunction from a limited number of
basic cellular processes common to eukaryotic cells was introduced and later
expanded on in a pair of essays by Hanahan and Weinberg [@weinberg2000;
@hallmarks_II]. An alternative hypothesis is that every tumor belongs to one of
a large number of syndromes which are unique to each tissue-of-origin, that
share some mechanisms and treatment strategies. Recent publication of TCGA
consortium papers present a view largely, and surprisingly, consistent with this
latter hypothesis [@hoadleyMultiplatformAnalysis2014;
@hoadleyCellofOriginPatternsDominate2018]. Perhaps owing to the intractable
complexity of genomics, proteomics, and patient metadata in all its forms, the
inescapable conclusion thus far is that tissue-of-origin remains the most
important driver of tumor characteristics at every scale and by every measure.
Our observations contrast with this view, and instead support an interpretation
of publicly available data in which all tumors manifest one of a limited number
of phenotypes resulting from disruptions of basic pathways.

We attempted to account for our clusters in terms of more trivial explanations.
For example, it could be that the clusters we identified are consistent with
disease progression. We were unable to identify any such trend in the number of
mutations, the relative staging or metastasis; and each cluster instead was
associated with unique combinations of pathways.

Some cancer types are unevenly distributed among the clusters, though we could
not identify any cancers that were exclusive to a single cluster. Only cancer
types with the fewest samples were found to be absent from one or more clusters
at all. Somewhat surprisingly to us, this finding extends to histological
subtypes of breast, head and neck cancers, leukemias, _etc._ This result implies
that histological subtypes could reflect differences in cell-of-origin, rather
than fundamental differences in cancer phenotype. The four major subtypes of
breast cancer correspond to histological and molecular expression profiles that
define them and how they respond to experimental stimuli [@breast-subtypes2]. It
has been hypothesized that differences in the molecular regulators of
development in the precursor cell types present in the breast epithelium drive
the various histological phenotypes [@breast-subtypes; @breast-subtypes3].
Consistent with this view we found that breast tumor samples of the Luminal A
subtype were heavily biased toward membership in clusters 1 and 7, and basal
tumor samples were biased toward cluster 6. However, both subtypes also
contained samples in every other cluster (without exception), and Luminal B and
Her2 positive samples are distributed across clusters. We interpret these data
to mean that inherited cell-of-origin signatures could predispose certain
precursor cells within the breast epithelium to forming tumors of one cluster or
another but are not determinative. This view is compatible with the previously
stated hypothesis but opens the way for a more granular view of individual
tumors. It would be surprising if we did not observe bias for some cancers and
subtypes amongst our classes, since some treatment regimens have greater
efficacy for patients of a given cancer or histological subtype
[@breast-subtypes2]. Nonetheless, the basis for some tumors being
treatment-refractory in spite of receiving the standard of clinical care for
diagnostic markers remains elusive. Doubtless some of this is due to chance
events, as tumors can metastasize and remain dormant years before they are
detected at distal sites, or resistant clones may have already arisen at
undetectable levels [@hallmarks_II], but our analysis suggests the possibility
of identifying more informative molecular, histological or cellular subtypes
that could form a basis for future stratification of patients into different
precision treatment regimens.

@@

Our results illustrate how unique combinations of mutations in pan-cancer driver
genes with tissue-specific pathway disruptions result in common categories when
viewed at the level of the pathway knowledge-base. Top cancer driver genes (_e.g.
PIK3CA_ and _TP53_) are found in most of the clusters, in spite of the fact that
they contribute to many cluster-specific pathways. This can only be explained as
a result of other less common driver genes complementing the pathway disruption
of the unique combination of driver genes that are disrupted in each tumor, and
we speculate that many of these are sensitive to mutation in different tissues.

@@

Finally, we must remark on the limitations of our work exemplified in two of the
clusters, 1 and 10, for which we did not find many distinctive associations with
pathways. Cluster 1 had a relatively low proportion of mutated pathways,
although it is broadly enriched in many of the same tumor-promoting pathways
common to the other groups. In addition, our data show clearly that this cluster
is as likely to contain stage IV metastatic tumors as it is to contain those of
stage I. This cluster likely represents a group of tumors with aberrations in
methylation, copy number, or other structural variants. Consistent with this,
kidney chromophobe and thyroid cancers have high proportions of structural
variations _vs._ other variant types [@PCAWG-MAIN] and are heavily skewed to
cluster 1 membership. Likewise, cluster 10 represents a group of hyper-mutated
tumors that harbor so many mutations that virtually no pathway is unaffected. It
seems likely that a significant fraction of the “mutant” samples for each
pathway are burdened with excess passenger mutations. This could be addressed
with more sophisticated filtering of likely passenger mutations [_e.g._
@spliceAI; @deep-learning-clinical-variants]. In the future, we hope to
incorporate these other data into a comprehensive pathway-centric analysis as we
have done here for point mutations and indels.

@@

By modeling CDC longevity data as a baseline risk function we showed that each
of our pathway disruption clusters exhibit cancer-type specific effects on
expected survival. However, considering that within each cancer type there are
different clinical standards of care, and even within classes of drugs the
preferred treatment can very between cancers, it makes sense that we observe
tissue-specific cluster effects. Contrast the situation with ovarian _vs._
breast cancer, which are both hormonally driven cancers, for example. Ovarian
cancer has but one main treatment axis, platinum, whereas breast cancer patients
have a variety of treatment regimens based on molecular subtype and other
factors. Unfortunately, given the diversity of drug classes and treatments, we
lack sufficient power to explore these variables in the TCGA data. It is our
hope that future studies will help to distinguish between treatment-specific
effects on survival given different pathway disruption clusters.

@@

Our findings imply that there are separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, cluster-specific
mutations and metastasis. General cancer promoting processes include genome
stability and immortality, as "enabling characteristics" of the cancer phenotype
[@hallmarks_II]. Such pathways are disrupted in most clusters and are frequently
the result of aberrations involving common driver genes such as BRCA1/2, MMR
genes, mitotic checkpoints, cohesion complexes, _etc._ Cluster-specific
evolution must involve the acquisition of disruptions to pathways that may
individually be harmful (_e.g._ highly proliferative cells are more likely to
senesce) but together produce more specialized cancer phenotype and increased
fitness. Importantly, our observations do not imply the order in which these
mutations should accumulate. This could be addressed in a future study by
evolutionary analysis of clonality, drawing inference from variant allele
frequencies as in @PCAWG-EVO. However, since many of the genes in the
non-cluster-specific pathways involve the known driver genes, it is reasonable
to surmise that these mutations promote or enable acquisition of
cluster-specific defects via random mutation and natural selection, thus
producing the clusters we observed. In support of this, the pan-cancer analysis
of whole genomes consortium (PCAWG) found that oncogenic driver mutations are
highly enriched in early arising clones, whereas later arising clones have much
greater diversity in driver mutations [@PCAWG-EVO]. Moreover, driver genes that
are known to be responsible for discrete mutation signatures such as _APOBEC_,
_BRCA1_ and _BRCA2_ produce mutational hotspots reflecting varying selective
pressures in different tissues [@PCAWG-MAIN].

>>>>

One of the biggest hurdles in cancer research is the sparsity of data; ~20,000
protein-coding genes is comparable with the number of tumor samples, even with
multiple mutations per sample. We sought to simplify the problem by employing a
"knowledge-base driven analysis" [@pathways-Atul-Butte], investigating cancer as
a disease of basic cellular and biochemical pathways. We accomplished this by
translating gene-level mutations into pathway level disruptions. Our approach
differs from previously described methods [@PATHNET] in that we chose to focus
on the pathway (defined in the methods) as the unit of disruption instead of the
gene, where individual mutations may be sufficient to alter pathway activity.

To our knowledge, this approach has not previously been attempted despite its
relative simplicity. We limited our analysis to mutations with likely
deleterious effects in genes that are actively expressed in each cancer type,
thus avoiding bias from transcription coupled repair. Our method of filtration
differs from the "rank-and-cut" method of [@PCAWG-MAIN] but represents a
reasonable attempt to account for the same biases. We also restricted our
analysis to biochemical pathways, excluding curated gene sets related to
diseases, syndromes, or classes of proteins with shared catalytic activity or
conserved domains which are potentially problematic [@pathways-Atul-Butte]. We
chose this approach to limit redundancy and exclude biologically unrelated
collections of genes.

The hypothesis that cancer results from dysfunction in basic cellular processes
common to eukaryotic cells was introduced and later expanded on in a pair of
essays by Hanahan and Weinberg [@weinberg2000;
@hallmarks_II]. An alternative hypothesis is that every tumor belongs to one of
a large number of syndromes which are unique to each tissue-of-origin, that
share some mechanisms and treatment strategies. Recent publication of TCGA
consortium papers present a view largely, and surprisingly, consistent with this
latter hypothesis [@hoadleyMultiplatformAnalysis2014;
@hoadleyCellofOriginPatternsDominate2018]. Perhaps owing to the intractable
complexity of genomics, proteomics, and patient metadata in all its forms, the
inescapable conclusion thus far is that tissue-of-origin remains the most
important driver of tumor characteristics at every scale and by every measure.
Our observations contrast with this view, and instead support an interpretation
of publicly available data in which all tumors manifest one of a limited number
of phenotypes resulting from disruptions of basic pathways.

We attempted to account for our clusters in terms of more trivial explanations.
For example, it could be that the clusters are consistent with disease
progression. We were unable to identify any such trend in the number of
mutations, the relative staging or metastasis; and each cluster instead was
associated with unique combinations of pathways. Some cancer types are unevenly
distributed among the clusters, though we could not identify any cancers that
were exclusive to a single cluster. Only cancer types with the fewest samples
were found to be absent from one or more clusters at all.

Somewhat surprisingly to us, this finding extends to histological subtypes of
breast, head and neck cancers, leukemias, _etc._ This result implies that
histological subtypes could reflect differences in cell-of-origin, rather than
fundamental differences in cancer phenotype. The four major subtypes of breast
cancer correspond to histological and molecular expression profiles that define
them and how they respond to experimental stimuli [@breast-subtypes2]. It has
been hypothesized that differences in molecular regulators of development in
precursor cell types present in breast epithelium drive histological phenotypes
[@breast-subtypes; @breast-subtypes3]. Consistent with this view we found that
breast tumor samples of the Luminal A subtype were heavily biased toward
membership in clusters 1 and 7, and basal tumor samples were biased toward
cluster 6. However, both subtypes also contained samples in every other cluster
(without exception), and Luminal B and Her2 positive samples are distributed
across clusters. Our interpretation of these data is that inherited
cell-of-origin signatures could predispose certain precursor cells within the
breast epithelium to forming tumors of one cluster or another but are not
determinative. This view is compatible with the previously stated hypothesis but
opens the way for a more granular view of individual tumors.

It would be surprising if we did not observe bias for some cancers and
subtypes amongst our classes, since some treatment regimens have greater
efficacy for patients of a given cancer or histological subtype
[@breast-subtypes2]. Nonetheless, the basis for some tumors being
treatment-refractory in spite of receiving the standard of clinical care for
diagnostic markers remains elusive. Doubtless some of this is due to chance
events, as tumors can metastasize and remain dormant years before they are
detected at distal sites, or resistant clones may have already arisen at
undetectable levels [@hallmarks_II], but our analysis suggests the possibility
of identifying more informative molecular, histological or cellular subtypes
that could form a basis for future stratification of patients into different
precision treatment regimens.

@@

Our results illustrate how unique combinations of mutations in pan-cancer driver
genes with tissue-specific pathway disruptions result in common categories when
viewed at the level of the pathway knowledge-base. Top cancer driver genes
(_e.g. PIK3CA_ and _TP53_) are found in most of the clusters, in spite of the
fact that they contribute to many cluster-specific pathways. This can only be
explained by less common genes complementing the unique combination of driver
genes in each tumor, and we speculate that many of these less common genes are
sensitive to increased mutation frequency in different tissues.

@@

Finally, we must remark on the limitations of our work exemplified in 
clusters 1 and 10, for which we did not find many distinctive associations with
pathways. Cluster 1 had a relatively low proportion of mutated pathways,
although it is broadly enriched in many of the same tumor-promoting pathways
common to other clusters. In addition, our data show clearly that this cluster
is as likely to contain stage IV metastatic tumors as it is to contain those of
stage I. This cluster likely represents a group of tumors with aberrations in
methylation, copy number, or other structural variants. Consistent with this,
kidney chromophobe and thyroid cancers have high proportions of structural
variations _vs._ other variant types [@PCAWG-MAIN] and are heavily skewed to
cluster 1 membership. Likewise, cluster 10 represents a group of hyper-mutated
tumors that harbor so many mutations that virtually no pathway is unaffected. It
seems likely that a significant fraction of the “mutant” samples for each
pathway are burdened with excess passenger mutations. This could be addressed
with more sophisticated filtering of likely passenger mutations [_e.g._
@spliceAI; @deep-learning-clinical-variants]. In the future, we hope to
incorporate these other data into a comprehensive pathway-centric analysis as we
have done here for point mutations and indels.

@@

By modeling CDC longevity data as a baseline risk function we showed that each
of cluster exhibits cancer-type specific effects on survival expectancy.
However, considering that within each cancer type there are different clinical
standards of care, and even within classes of drugs the preferred treatment can
very between cancers, it makes sense that we observe tissue-specific cluster
effects. Contrast ovarian _vs._ breast cancer, which are both hormonally driven
cancers, for example. Ovarian cancer has but one main treatment axis, platinum,
whereas breast cancer patients have a variety of treatment regimens based on
molecular subtype and other factors. Unfortunately, given the diversity of drug
classes and treatments, we lack sufficient power to explore these variables in
the TCGA data. It is our hope that future studies will help to distinguish
between treatment-specific effects on survival given different pathway
disruption clusters.

@@

Our findings imply separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, cluster-specific
mutations and metastasis. General cancer promoting processes include genome
stability and immortality, as "enabling characteristics" of the cancer phenotype
[@hallmarks_II]. Such pathways are disrupted in most clusters and are frequently
the result of aberrations involving common driver genes such as BRCA1/2, MMR
genes, mitotic checkpoints, cohesion complexes, _etc._ Cluster-specific
evolution must involve the acquisition of disruptions to pathways that may
individually be harmful (_e.g._ highly proliferative cells are more likely to
senesce) but together produce more specialized cancer phenotype and increased
fitness. Importantly, our observations do not imply the order in which these
mutations should accumulate. This could be addressed in a future study by
evolutionary analysis of clonality, drawing inference from variant allele
frequencies as in @PCAWG-EVO. However, since many of the genes in the
non-cluster-specific pathways involve the known driver genes, it is reasonable
to surmise that these mutations promote or enable acquisition of
cluster-specific defects via random mutation and natural selection, thus
producing the clusters we observed. In support of this, the pan-cancer analysis
of whole genomes consortium (PCAWG) found that oncogenic driver mutations are
highly enriched in early arising clones, whereas later arising clones have much
greater diversity in driver mutations [@PCAWG-EVO]. Moreover, driver genes that
are known to be responsible for discrete mutation signatures such as _APOBEC_,
_BRCA1_ and _BRCA2_ produce mutational hotspots reflecting varying selective
pressures in different tissues [@PCAWG-MAIN].

### 2020-09-12 text/manuscript.Rmd

To study cancer pathways we obtained a set of 7,607 solid tumors from The Cancer
Genome Atlas (TCGA) through the [Genomic Data Commons (GDC)](gdc.cancer.gov)
portal. TCGA data are most appropriate given the relative completeness of the
patient metadata for survival and staging. We chose to analyze exome sequencing
data because the affected target gene is known unambiguously. Therefore, we
selected all missense, nonsense, frameshift, stop-loss, untranslated region, and
splicing mutations. To minimize bias from well-studied diseases and processes,
we selected 377 [Reactome pathways](https://reactome.org/) (see supplemental
table ST1) of interest corresponding to basic cellular processes and biochemical
pathways, excluding those corresponding to catalytic categories (_e.g._
“transcription factors”) or disease associations (_e.g._ “mutated in colon
cancer”) and filtered our gene list on membership in these pathways (total of
8,940 genes).

@@

We investigated this dataset using multiple correspondence analysis (MCA)
[@JSSv025i01], and visually summarized the analysis with UMAP (**figure 1B** and
see interactive media from [**supplemental file
1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)) [@McInnes2018]. We
used the resulting graph coordinates to perform density based clustering
with HDBSCAN [@hdbscan], which resulted in identification of 10 well-defined
clusters capturing about 80% of the tumor samples. To capture the remaining
samples into one of these 10 clusters we used kNN (see **Supplementary Methods**
for details on clustering methods).

@@

Many cancers have molecular or histological subtypes defined based on gene
expression, pathology or other -omics profiles. These subtypes often have
different standards of care owing to different drug sensitivity (or other
clinical trial data). If histological subtypes represent true molecular
phenotypes, one predicts they should segregate with pathway-based clusters,
supporting the clusters as proxies for molecular pathology sub-typing. To our
surprise, we find a similar result to the previous analysis of cancer types. To
illustrate this, we projected annotations for each of the breast cancer
subtypes, composed of Triple-negative/Basal-like, Her2 positive, normal-like,
and luminal A and B subtypes onto the UMAP. These are among the most heavily
studied molecular subtypes in cancer, each with different prognoses and
standards of care. We did not observe exclusive segregation by cluster for these
subtypes (**figure 2B**). We also projected histological subtype data for the
remaining cancers (see **supplementary figure S2** and online [**supplemental
file1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)); we find that
the subtypes, though often biased to one or more clusters, are almost never
exclusive. We interpret these data to mean that the clusters do not correspond
to previously identified molecular subtypes within the parent cancer types.

@@

There are several familial cancer-causing mutations studied for differences in
basic biology, survival and treatment outcomes. The functions of these genes are
related to risk factors such as genome stability, proof-reading & DNA damage
repair, and telomere length. _BRCA1/2_ genes are key for DNA double-stranded
break repair [@moynihan-1999-brca1;@davies-2001-brca2] and germline mutations in
these genes confer elevated risk for breast, prostate and ovarian cancers. The
mechanism of risk is thought to involve loss of heterozygosity, resulting in
loss of the wildtype, functional allele [@brca-loh], so we projected somatic
mutations for _BRCA1_ and _BRCA2_ genes onto the UMAP. We did not observe
segregation of these mutations into specific clusters (**figure 2C** and see
interactive media from [**supplemental file
1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). We also projected
mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_, _MLH3_, _PMS1_ and _PMS2_;
_BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these genes except for _APC_
exhibited remarkable specificity with respect to cluster assignment
(**supplementary figure S3**). To look at other risk factors such as maintenance
of DNA methylation levels and telomere length, we projected somatic mutations of
the _TET2_ and _TET3_ genes, plus _TERT_, _TEP1_, and _DKC1_, and observed
similar lack of cluster bias (**supplementary figure S3**).

@@

To discover what pathways are most important for clustering, we calculated
percent enrichment _within cluster_ relative to all other clusters and ranked
pathways from highest to lowest enrichment. We visualized enrichment as a
heatmap (**figure 3A**). Using this approach, we identified about fourteen
pathways per cluster (enrichment score ≥ 0.3, 95% confidence; see methods)
(**supplementary table ST2**). Clusters 7, 8, and 9 had several pathways in
common. To explore the specific pathways marking each cluster, we projected
disruptions for each of the 377 pathways onto the UMAP (**supplementary table
ST2** and see interactive media from [**supplemental file
1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Clusters 3 and 5
were distinguished by metabolic pathways including RNA and protein biosynthesis
(**supplementary table ST2**). Similarly, cluster 4 was distinguished by
mutations affecting regulation of DNA and histone methylation ("DNA
methylation", "PRC2 methylates histones and DNA", and "Nucleosome assembly").
Clusters 7-9 have in common mutations in extracellular, intracellular, and
immune-related signaling pathways (see **figure 3B** and **supplementary table
ST2**). Cluster 2 had the highest pathway enrichment levels of the three, having
mutations in hedgehog signaling, "β-catenin degradation", "cellular response to
hypoxia", "regulation of cell cycle" and "apoptosis" among others.

@@

Prior efforts to extract signatures from pan-cancer datasets met with difficulty
in distinguishing tumor samples from tissue-specific -omics data signatures.
Given our pathway-disruption based clustering, this raises the question, are
tumor phenotypes driven by common driver genes, “silent” tissue-specific
effectors (_i.e._ too few samples to detect above statistical significance
thresholds), or a combination of both? To answer this question, we compared top
pathway genes for each cluster relative to TCGA background to find
differentially mutated genes. We ranked odds ratios and selected the top ten
enriched and depleted genes (pvalue < 0.01) for each cluster (**figure 4**; odds
ratios plot). Clusters 7 and 8, which shared multiple enrichment in signaling
pathways, are largely driven by mutations in _PI3K_ and its orthologs and _Ras_
genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **Supplementary
figure S4** and see interactive media from [**supplemental file
1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Interestingly,
cluster 9, which also shared multiple enrichment in signaling pathways with
clusters 7 and 8, is enriched for both _PIK3CA_ and _KRAS_. Clusters 3 and 5,
defined by enrichment in metabolic pathways, had mutations in ribosomal proteins
and nuclear pore complex, respectively. Cluster 4 had mutations nucleosome
structural or subunit genes. Cluster 2 had mutations in proteasomal subunits
involved in protein degradation. We also observed that genes enriched for one
cluster are depleted from others (_i.e._ is enriched in cluster 6, but depleted
in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in clusters 3 and
8). Next, we investigated the proportion of samples per cancer type for the
significant genes within a cluster (**figure 4**; heatmap). Surprisingly,
clusters were not predominated by one or more highly mutated genes across all
cancers. Instead, when observing the mutation rate for these genes within
samples that belong to a cluster, the mutation rate is heterogeneous across
tumors by tissue origin (_e.g._ In cluster 4, CESC was enriched for _H2AFX_, OV
was enriched for _HIST1H2BD_, and UCEC was enriched for _HIST1H2AC_). Even among
the top enriched genes within clusters there is no global pattern, indicating
that our clusters are not driven by individual genes, but rather networks as a
whole. Taken together, our data identify a framework of cancer type-specific
mutations associated with specific clusters.

@@

All code and data for producing the analyses and figures herein are included in
this fully reproducible manuscript in R markdown format. R markdown files and
all other models are available from [our
repositories](https://github.com/dennishazelett/TTmanu) on the distributed
version control site, GitHub. Extended descriptions of clustering, models,
formulae, pathway selection, and gene filtering are provided in supplementary
methods.

>>>>

To study cancer pathways we obtained a set of 7,607 solid tumors from The Cancer
Genome Atlas (TCGA) through the Genomic Data Commons (GDC)(gdc.cancer.gov)
portal. TCGA data are most appropriate given the relative completeness of the
patient metadata for survival and staging. We chose to analyze exome sequencing
data because the affected target gene is known unambiguously. Therefore, we
selected all missense, nonsense, frameshift, stop-loss, untranslated region, and
splicing mutations. To minimize bias from well-studied diseases and processes,
we selected 377 Reactome pathways (reactome.org) (see supplemental
table ST1) of interest corresponding to basic cellular processes and biochemical
pathways, excluding those corresponding to catalytic categories (_e.g._
“transcription factors”) or disease associations (_e.g._ “mutated in colon
cancer”) and filtered our gene list on membership in these pathways (total of
8,940 genes).

@@

We investigated this dataset using multiple correspondence analysis (MCA)
[@JSSv025i01], and visually summarized the analysis with UMAP (**figure 1B** and
see interactive media from **supplemental file
1** (junkdnalab.shinyapps.io/PANCAN_supplemental/)) [@McInnes2018]. We
used the resulting graph coordinates to perform density based clustering
with HDBSCAN [@hdbscan], which resulted in identification of 10 well-defined
clusters capturing about 80% of the tumor samples. To capture the remaining
samples into one of these 10 clusters we used kNN (see **Supplementary Methods**
for details on clustering methods).

@@

Many cancers have molecular or histological subtypes defined based on gene
expression, pathology or other -omics profiles. These subtypes often have
different standards of care owing to different drug sensitivity (or other
clinical trial data). If histological subtypes represent true molecular
phenotypes, one predicts they should segregate with pathway-based clusters,
supporting the clusters as proxies for molecular pathology sub-typing. To our
surprise, we find a similar result to the previous analysis of cancer types. To
illustrate this, we projected annotations for each of the breast cancer
subtypes, composed of Triple-negative/Basal-like, Her2 positive, normal-like,
and luminal A and B subtypes onto the UMAP. These are among the most heavily
studied molecular subtypes in cancer, each with different prognoses and
standards of care. We did not observe exclusive segregation by cluster for these
subtypes (**figure 2B**). We also projected histological subtype data for the
remaining cancers (see **supplementary figure S2** and see interactive media from 
**supplemental file 1**); we find that
the subtypes, though often biased to one or more clusters, are almost never
exclusive. We interpret these data to mean that the clusters do not correspond
to previously identified molecular subtypes within the parent cancer types.

@@

There are several familial cancer-causing mutations studied for differences in
basic biology, survival and treatment outcomes. The functions of these genes are
related to risk factors such as genome stability, proof-reading & DNA damage
repair, and telomere length. _BRCA1/2_ genes are key for DNA double-stranded
break repair [@moynihan-1999-brca1;@davies-2001-brca2] and germline mutations in
these genes confer elevated risk for breast, prostate and ovarian cancers. The
mechanism of risk is thought to involve loss of heterozygosity, resulting in
loss of the wildtype, functional allele [@brca-loh], so we projected somatic
mutations for _BRCA1_ and _BRCA2_ genes onto the UMAP. We did not observe
segregation of these mutations into specific clusters (**figure 2C** and see
interactive media from **supplemental file 1**). We also projected
mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_, _MLH3_, _PMS1_ and _PMS2_;
_BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these genes except for _APC_
exhibited remarkable specificity with respect to cluster assignment
(**supplementary figure S3**). To look at other risk factors such as maintenance
of DNA methylation levels and telomere length, we projected somatic mutations of
the _TET2_ and _TET3_ genes, plus _TERT_, _TEP1_, and _DKC1_, and observed
similar lack of cluster bias (**supplementary figure S3**).

@@

To discover what pathways are most important for clustering, we calculated
percent enrichment _within cluster_ relative to all other clusters and ranked
pathways from highest to lowest enrichment. We visualized enrichment as a
heatmap (**figure 3A**). Using this approach, we identified about fourteen
pathways per cluster (enrichment score ≥ 0.3, 95% confidence; see methods)
(**supplementary table ST2**). Clusters 7, 8, and 9 had several pathways in
common. To explore the specific pathways marking each cluster, we projected
disruptions for each of the 377 pathways onto the UMAP (**supplementary table
ST2** and see interactive media from **supplemental file
1**). Clusters 3 and 5
were distinguished by metabolic pathways including RNA and protein biosynthesis
(**supplementary table ST2**). Similarly, cluster 4 was distinguished by
mutations affecting regulation of DNA and histone methylation ("DNA
methylation", "PRC2 methylates histones and DNA", and "Nucleosome assembly").
Clusters 7-9 have in common mutations in extracellular, intracellular, and
immune-related signaling pathways (see **figure 3B** and **supplementary table
ST2**). Cluster 2 had the highest pathway enrichment levels of the three, having
mutations in hedgehog signaling, "β-catenin degradation", "cellular response to
hypoxia", "regulation of cell cycle" and "apoptosis" among others.

@@

Prior efforts to extract signatures from pan-cancer datasets met with difficulty
in distinguishing tumor samples from tissue-specific -omics data signatures.
Given our pathway-disruption based clustering, this raises the question, are
tumor phenotypes driven by common driver genes, “silent” tissue-specific
effectors (_i.e._ too few samples to detect above statistical significance
thresholds), or a combination of both? To answer this question, we compared top
pathway genes for each cluster relative to TCGA background to find
differentially mutated genes. We ranked odds ratios and selected the top ten
enriched and depleted genes (pvalue < 0.01) for each cluster (**figure 4**; odds
ratios plot). Clusters 7 and 8, which shared multiple enrichment in signaling
pathways, are largely driven by mutations in _PI3K_ and its orthologs and _Ras_
genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **Supplementary
figure S4** and see interactive media from **supplemental file 1**. Interestingly,
cluster 9, which also shared multiple enrichment in signaling pathways with
clusters 7 and 8, is enriched for both _PIK3CA_ and _KRAS_. Clusters 3 and 5,
defined by enrichment in metabolic pathways, had mutations in ribosomal proteins
and nuclear pore complex, respectively. Cluster 4 had mutations nucleosome
structural or subunit genes. Cluster 2 had mutations in proteasomal subunits
involved in protein degradation. We also observed that genes enriched for one
cluster are depleted from others (_i.e._ is enriched in cluster 6, but depleted
in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in clusters 3 and
8). Next, we investigated the proportion of samples per cancer type for the
significant genes within a cluster (**figure 4**; heatmap). Surprisingly,
clusters were not predominated by one or more highly mutated genes across all
cancers. Instead, when observing the mutation rate for these genes within
samples that belong to a cluster, the mutation rate is heterogeneous across
tumors by tissue origin (_e.g._ In cluster 4, CESC was enriched for _H2AFX_, OV
was enriched for _HIST1H2BD_, and UCEC was enriched for _HIST1H2AC_). Even among
the top enriched genes within clusters there is no global pattern, indicating
that our clusters are not driven by individual genes, but rather networks as a
whole. Taken together, our data identify a framework of cancer type-specific
mutations associated with specific clusters.

@@

All code and data for producing the analyses and figures herein are included in
this fully reproducible manuscript in R markdown format. R markdown files and
all other models are available from our
repositories (https://github.com/dennishazelett/TTmanu) on the distributed
version control site, GitHub. Extended descriptions of clustering, models,
formulae, pathway selection, and gene filtering are provided in supplementary
methods. An interactive media of the results is available on junkdnalab.shinyapps.io/PANCAN_supplemental/.
